---
title: "plotting evolutionary events related to genes of specific substrates"
author: "Jimmy Zhong"
date: "5/5/2023"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(cowplot)
library(dplyr)
library(stringr)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path ))

clock_model = "ugam1"
categories_COGs = read.csv("high-cor-COGs.txt", row.names=1, comment.char="#")
COG_record = read.csv(sprintf("../COG_reconciliation_summary_%s.csv", clock_model))
done_COGs = pull(COG_record, COG)
good_COGs = filter(COG_record, exit_status == "everything_successful") %>% pull(COG)
```

## exploration
Switched between use left.date, midpoint.date, and right.date.

Right.date is bad: everything got squished to the right.
Left.date gives a large spike around 1500 million years ago. Looks like bioinformatics artifacts.
Midpoint.date: move duplication, horizontal gene transfer, and gene lost events to the right, but speciation events stays the same place. Resulting in the unique shape we see -- the speciation events are more dated back!

```{r}
get_COGs_of = function(key) {
  COGs = str_split(categories_COGs[key,], " ")[[1]]
  bad_COGs = setdiff(COGs, good_COGs)
  print(paste("I don't have these COGs", paste(bad_COGs, collapse = " ")))
  print("These COG failed:")
  fail_COGs = intersect(bad_COGs, done_COGs)
  print(COG_record %>% filter(COG %in% fail_COGs) %>% select(COG, exit_status) )
  print(paste("I have not run to these COGs yet:", paste(setdiff(bad_COGs, done_COGs), collapse = " ")))
  good_COGs = intersect(COGs, good_COGs)
  print(paste("I have these good COGs for :", key))
  print(paste(good_COGs, collapse = " "))
  return(good_COGs)
}

read_events = function(file) {
  all = read.delim(file, na.strings = "?")
  COG = str_match(file, "COG\\d+")[1]
  all %>% filter(right.node != "UNSAMPLED") %>%
    select(event, left.date, midpoint.date, right.date) %>%
    mutate(COG = COG)
}

merge_events = function(COGs) {
  file_list = paste0(sprintf("../%s_ecceTERA_analysis/", clock_model),
                     COGs, "_symmetric.events_event_dates.txt")
  data_list = lapply(file_list, read_events)
  merge_data = do.call(rbind, mapply(rbind, data_list, SIMPLIFY = FALSE))
}

shared_gen_graph = function(data) {
  g = ggplot(data, aes(x=right.date)) + # midpoint.date
    geom_histogram(aes(y=(..density..), fill=event), binwidth = 50, boundary = 0)
  if (type == "density") {
    g = ggplot(data, aes(x=left.date)) + # midpoint.date
      geom_density(alpha=.2, aes(fill=event))
  }
  g + xlim(4000,0) +
    ylab("Proportion of total events") +
    theme(legend.position = "bottom",
  		axis.line.y = element_line(linetype = 1)) + 
  	geom_hline(yintercept = 0) +
  	theme_classic()
}

gen_graph = function(events, plot_title = "merged events"){
  top = shared_gen_graph(filter(events, event %in% c("dup", "spe"))) +
    theme(plot.title = element_text(size = rel(1.75), hjust = 0.5),
      axis.line.x = element_blank(),
      axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      axis.ticks.x = element_blank()) +
    labs(title=plot_title) + 
    scale_fill_manual(values = c("dup" = "#06d6a0", "spe" = "#ff66de"))
  bottom = shared_gen_graph(filter(events, event %in% c("hgt", "los"))) + 
    scale_y_continuous(trans = "reverse") + 
    scale_fill_manual( values = c("hgt" = "#ffd166", "los" = "#118ab2"))
  plot_grid(top, bottom, ncol = 1, align = "v", axis = "tb")
}

get_events = function(feature, anno) {
  COGs = get_COGs_of(feature)
  events = merge_events(COGs)
  events$substrate = anno
  return(events)
}

make_graph = function(events, out_graph, title_text) {
  num_COGs = length(unique(events$COG))
  pt = sprintf("%s (n = %d)", title_text, num_COGs)
  gen_graph(events, plot_title = pt)
  ggsave(out_graph, width = 6, height = 5)
}
```

```{r}
iron = get_events("sqrt_iron", "iron")
Fe2_iron = get_events("Fe2+_text_search", "Fe2_textSearch")
Fe3_iron = get_events("Fe3+_text_search", "Fe3_textSearch")

NO2NO3 = get_events("log_NO2NO3", "Nitrogen")
PO4 = get_events("log_PO4", "PO4")
oxygen = get_events("Mean_Oxygen", "Oxygen")
inorganic_ion = get_events("inorganic_ion", "inorganic_ion")

# type="density"
type="histogram"

gen_name = function(key_word) return(sprintf("%s-plots/right_merge_events_%s_%s_%s.png", clock_model, key_word, clock_model, type))

make_graph(iron, gen_name("iron"), "COGs high corr to iron")
make_graph(NO2NO3, gen_name("nitrogen"), "COGs high corr to nitrogen")
make_graph(PO4, gen_name("phosphate"), "COGs high corr to phosphate")
make_graph(oxygen, gen_name("oxygen"), "COGs high corr to oxygen")
make_graph(inorganic_ion, gen_name("inorganic_ion"), "inorganic ion COGs")
```

## plotting everything together
```{r}
all_dfs = rbind(iron, Fe2_iron, Fe3_iron, NO2NO3, PO4, oxygen, inorganic_ion)
# all_dfs = all_dfs %>% mutate(substrate = fct_relevel(substrate, c("")))

ggplot(all_dfs, aes(x=midpoint.date, y=substrate, color = event, group = substrate)) + # midpoint.date
  geom_jitter(alpha = 0.3, height = 0.3) +
    scale_x_reverse(limits = c(4000, 0),
                    breaks = seq(0, 4000, by = 500)) +
    scale_color_manual(values = c("dup" = "#06d6a0",
                                  "hgt" = "#ffd166",
                                  "los" = "#118ab2",
                                  "spe" = "#ff66de")) +
    labs(x = "Million years ago", y = "") +
    # geom_hline(yintercept = 20.5) +
    # geom_hline(yintercept = 19.5, color = "lightgray") +
  annotate('rect', xmin=2500, xmax=2300, ymin=0, ymax=length(unique(all_dfs$substrate)) + 1, alpha=.3, fill='gray') +
    theme(panel.background = element_blank(),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.line = element_line(colour = "black"))

ggsave("all-substrate-jitter.png", width = 10, height = 5)
```