---
title: "plotting evolutionary events related to genes of specific substrates"
author: "Jimmy Zhong"
date: "5/5/2023"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(cowplot)
library(dplyr)
library(stringr)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path ))
source("plot-substrate-utils.R")
```

## set important parameters!!!
```{r}
clock_model = "ugam1"
remove_leaf_events = TRUE
plot_node = "midpoint"
# plot_node = "left"
# plot_node = "right"

# type="density"
type="histogram"
```

```{r}
categories_COGs = read.csv("high-cor-COGs.txt", row.names=1, comment.char="#")
COG_record = read.csv(sprintf("../COG_reconciliation_summary_%s.csv", clock_model))
done_COGs = pull(COG_record, COG)
good_COGs = filter(COG_record, exit_status == "everything_successful") %>% pull(COG)
```

## exploration
Switched between use left.date, midpoint.date, and right.date.

Right.date is bad: everything got squished to the right.
Left.date gives a large spike around 1500 million years ago. Looks like bioinformatics artifacts.
Midpoint.date: move duplication, horizontal gene transfer, and gene lost events to the right, but speciation events stays the same place. Resulting in the unique shape we see -- the speciation events are more dated back!

```{r}
iron = get_events("sqrt_iron", "iron")
Fe2_iron = get_events("Fe2+_text_search", "Fe2_textSearch")
Fe3_iron = get_events("Fe3+_text_search", "Fe3_textSearch")

NO2NO3 = get_events("log_NO2NO3", "Nitrogen")
PO4 = get_events("log_PO4", "PO4")
oxygen = get_events("Mean_Oxygen", "Oxygen")
inorganic_ion = get_events("inorganic_ion", "inorganic_ion")

# gen_name = function(k) return(sprintf("%s-plots/right_merge_events_%s_%s_%s.png", clock_model, k, clock_model, type))

gen_name = function(k) return(sprintf("internal-events-%snode-%s-plots/internal_%s_%s_%s.png", plot_node, clock_model, plot_node, k, type))

make_graph(iron, gen_name("iron"), "COGs high corr to iron")
make_graph(NO2NO3, gen_name("nitrogen"), "COGs high corr to nitrogen")
make_graph(PO4, gen_name("phosphate"), "COGs high corr to phosphate")
make_graph(oxygen, gen_name("oxygen"), "COGs high corr to oxygen")
make_graph(inorganic_ion, gen_name("inorganic_ion"), "inorganic ion COGs")
```

## plotting everything together
```{r}
all_dfs = rbind(iron, Fe2_iron, Fe3_iron, NO2NO3, PO4, oxygen, inorganic_ion)
# all_dfs = all_dfs %>% mutate(substrate = fct_relevel(substrate, c("")))

g1 = ggplot(all_dfs, aes(x=midpoint.date, y=substrate, color = event, group = substrate)) +
    scale_color_manual(values = c("dup" = "#06d6a0",
                                  "hgt" = "#ffd166",
                                  "los" = "#118ab2",
                                  "spe" = "#ff66de"))
jitter_plot_shared(g1, ytop = length(unique(all_dfs$substrate)))

ggsave(sprintf("%s-plots/all-substrate-jitter.png", clock_model), width = 10, height = 5)
```

## making histograms for locations 
```{r}
# can only choose left or right origin
plot_node = "left"
# plot_node = "right"

loc_iron = get_events("sqrt_iron", "iron", location = TRUE)
loc_NO2NO3 = get_events("log_NO2NO3", "Nitrogen", location = TRUE)
loc_PO4 = get_events("log_PO4", "PO4", location = TRUE)
loc_oxygen = get_events("Mean_Oxygen", "Oxygen", location = TRUE)
loc_inorganic_ion = get_events("inorganic_ion", "inorganic_ion", location = TRUE)

loc_dfs = rbind(loc_iron, loc_NO2NO3, loc_PO4, loc_oxygen, loc_inorganic_ion, get_events("Fe2+_text_search", "Fe2_textSearch", location = TRUE), get_events("Fe3+_text_search", "Fe3_textSearch", location = TRUE))

loc_dfs = loc_dfs[loc_dfs[, paste0(plot_node, ".origin")] %in% mixed_states, ]
```

```{r}
gen_loc_name = function(k) return(sprintf("internal-location-%snode-%s-plots/internal_%s_%s_%s.png", plot_node, clock_model, plot_node, k, type))

make_graph(loc_iron, out_graph = gen_loc_name("iron"), title_text = "COGs high corr to iron", location = TRUE)
make_graph(loc_NO2NO3, gen_loc_name("nitrogen"), "COGs high corr to NO2NO3", location = TRUE)
make_graph(loc_PO4, gen_loc_name("phosphate"), "COGs high corr to phosphate", location = TRUE)
make_graph(loc_oxygen, gen_loc_name("oxygen"), "COGs high corr to oxygen", location = TRUE)
make_graph(loc_inorganic_ion, gen_loc_name("inorganic_ion"), "inorganic ion COGs", location = TRUE)
```

## plotting the location where the events happened!!!!
```{r}
g2 = ggplot(loc_dfs, aes_string(x = paste0(plot_node, ".date"), color = paste0(plot_node, ".origin"), y = "substrate", group = "substrate")) + scale_color_manual(values = mixed_color_scale)

jitter_plot_shared(g2, ytop = length(unique(all_location_dfs$substrate)))

if (remove_leaf_events) {
  ggsave(sprintf("all-substrate-%s-removeLeaf.png", plot_node), width = 10, height = 5)
} else {
  ggsave(sprintf("all-substrate-%s-location.png", plot_node), width = 10, height = 5)
}
```

## What if we spearate the COGs with positive and negative correlations to the substrates
```{r}
# can only choose left or right origin
plot_node = "left"
# plot_node = "right"
loc_pos_oxygen = get_events("positive_correlation_oxygen", "positive_correlation_oxygen", location = TRUE)
loc_neg_oxygen = get_events("negative_correlation_oxygen", "negative_correlation_oxygen", location = TRUE)
loc_pos_NO2NO3 = get_events("positive_correlation_NO2NO3", "positive_correlation_NO2NO3", location = TRUE)
loc_neg_NO2NO3 = get_events("negative_correlation_NO2NO3", "negative_correlation_NO2NO3", location = TRUE)
loc_pos_iron = get_events("positive_correlation_iron", "positive_correlation_iron", location = TRUE)
loc_neg_iron = get_events("negative_correlation_iron", "negative_correlation_iron", location = TRUE)
loc_pos_phosphate = get_events("positive_correlation_phosphate", "positive_correlation_phosphate", location = TRUE)
loc_neg_phosphate = get_events("negative_correlation_phosphate", "negative_correlation_phosphate", location = TRUE)
# no correlation stuff for the inorganic ion

gen_name3 = function(key_word) return(sprintf("posNeg-location-%s-plots/location-%s-%s.png", clock_model, key_word, type))

if (remove_leaf_events) {
  gen_name3 = function(key_word) return(sprintf("internal-posNeg-location-%s-plots/internal-location-%s-%s.png", clock_model, key_word, type))
}

make_graph(loc_pos_oxygen, gen_name3("oxygen-pos"), "COGs positive corr to oxygen", location = TRUE)
make_graph(loc_neg_oxygen, gen_name3("oxygen-neg"), "COGs negative corr to oxygen", location = TRUE)
make_graph(loc_pos_NO2NO3, gen_name3("NO2NO3-pos"), "COGs positive corr to NO2NO3", location = TRUE)
make_graph(loc_neg_NO2NO3, gen_name3("NO2NO3-neg"), "COGs negative corr to NO2NO3", location = TRUE)
make_graph(loc_pos_iron, gen_name3("iron-pos"), "COGs positive corr to iron", location = TRUE)
make_graph(loc_neg_iron, gen_name3("iron-neg"), "COGs negative corr to iron", location = TRUE)
make_graph(loc_pos_phosphate, gen_name3("phosphate-pos"), "COGs positive corr to phosphate", location = TRUE)
make_graph(loc_neg_phosphate, gen_name3("phosphate-neg"), "COGs negative corr to phosphate", location = TRUE)
```